Parity games in pushdown systems with parameters

### 2023-06-12 reachability/article.tex
up git p

\subsection{Games}

% Games represent a model for interaction between decision makers. We study so-called turn-based games in extensive form with perfect information where the players have to make decisions again and again.
\iffalse
After its foundation by von Neumann and Morgenstern [vNM44] in the
1940s, game theory has quickly become a valuable tool for modelling the
interaction of several agents with individual and possibly conflicting objec-
tives. It has applications in many scientific fields such as economy, sociology,
biology, logic and computer science. Whereas the games used in the former
three areas are usually finite in the sense that any play of these games ends
after a finite number of steps (in fact often after only a single step), the
games used in logic and computer science are in general infinite.

@article{copeland1945john,
  title={John von Neumann and Oskar Morgenstern, theory of games and economic behavior},
  author={Copeland, Arthur H},
  journal={Bulletin of the American Mathematical Society},
  volume={51},
  number={7},
  pages={498--504},
  year={1945},
  publisher={American Mathematical Society}
}

More precisely, an infinite game is played for an infinite number of
rounds. In each round, one player chooses an action depending on the se-
quence of actions chosen in previous rounds. After infinitely many rounds,
an infinite sequence of actions has emerged, called the outcome of the game.
Each player receives a payoff determined by a payoff function mapping each
possible outcome to a real number in the interval [0, 1]. We will concen-
trate on games where the possible payoffs are just 0 and 1, in which case a
winning condition, given as an abstract set of possible outcomes, is used to
determine whether a player wins (payoff 0) or loses (payoff 1) an outcome.
In logic, games are used to evaluate a logical formula in a mathematical
structure and to compare mathematical structures with respect to their
(in)distinguishability by a certain logic. Games of the former kind are called
model checking games; games of the latter kind are called model comparison
games. Whereas model checking and model comparison games for plain
modal or first-order logic are finite games, infinite games arise as model
checking games for fixed point logics and as model comparison games for
infinitary logics.
In computer science, games are used for the verification and synthesis
of reactive systems, i.e. systems that interact with their environment. As
soon as the systems under consideration do not necessarily terminate, these
games become infinite. For example, the behaviour of an operating system
can be understood as an infinite game between the system and its users.

So far, the games used in logic and computer science have usually been
two-player zero-sum games, i.e. games with only two players where one
player wins if and only if the other player loses. For example, in a model
checking game, one player wants to show that the given formula holds in the
given structure whereas the other player wants to show that this is not the
case. Analogously, in a game used for verification, one player wants to show
that the given system is able to react against its environment in such a way
that the resulting execution path of the system fulfils the given specification
whereas the other player wants to show that this is not the case.

The research on infinite games in theoretical computer science is based on a
mixture of several motivations:
1. The beautiful classical theory of infinite games, as developed in descriptive
set theory, lacks an algorithmic content. Such an algorithmic orientation is
provided by the automata theoretic approach to infinite games, originating
in work of Church, B ̈uchi, McNaughton, and Rabin about fourty years ago.
2. Determinacy results for infinite games are closely related to complementation
results for logics and automata; thus, infinite games help to analyze logical
theories (the most prominent being the monadic theory S2S of two successor
functions, see e.g. [Th97]).
3. Games are a natural model of reactive computation, and infinite games are
thus a faithful representation of nonterminating reactive systems (for which
control problems can be solved in terms of providing winning strategies).
4. The model checking problem for logics like the μ-calculus can be formulated
as the question to determine the winner of an infinite game.
The purpose of this tutorial is to explain the core of the algorithmic (and
automata theoretic) theory of infinite games. In the present extended abstract,
only some key notions are explained, without technical details and proofs. In the
tutorial, more will be said about topics of current research as listed in Section 6
of this abstract.

@incollection{thomas1997languages,
  title={Languages, automata, and logic},
  author={Thomas, Wolfgang},
  booktitle={Handbook of formal languages},
  pages={389--455},
  year={1997},
  publisher={Springer}
}

A game is composed of an arena and a winning condition.
\iffalse
\mh{Two-player games oﬀer a very convenient framework to represent
interaction of a program with some (possibly hostile) environment. In this approach, the ﬁrst player represents the program while the second player simulates
the environment. A winning strategy expresses a property that must hold whatever the environment does, thus ﬁnding a winning strategy for the ﬁrst player
allows to synthesize a controller that restricts the program and ensures that the
property expressed by the winning condition always holds \cite{arnold2003games}.}
\fi
We will ﬁrst study
arenas and
 then introduce common winning conditions that will be of interest to us.

 \subsubsection{Arenas}

\begin{samepage}
An {\em arena} is a tuple $A=(S_0, S_1,{\rightarrow})$  which is composed of
\begin{itemize}
	\item two  disjoint  sets of configurations, $S_0$ and $S_1$, %respectively associated with player $0$ and $1$, 
	whose disjoint union $S_0 \cup S_1$ we denote by $S$, and
	\item a relation of %labeled 
		transitions
	 ${\rightarrow} \subseteq S \times %\Lambda \times 
	 				S$. \\
\end{itemize}
\end{samepage}

@@

 \subsubsection{Plays}

\noindent
%
%\noindent
A {\em play} in an arena $A=(S_0,S_1,\rightarrow)$ is a path in $(S,{\rightarrow})$ that is {\em maximal} in the following sense: it is either infinite or finite and if it is finite, then the target configuration is a dead end. 
A {\em partial play} in an arena $A=(S_0,S_1,\rightarrow)$ is a 
% sequence of configurations $s_{init} = s_0 s_1 \ldots \in S^\infty$ connected by the transition relation, i.e. such that $(v_i, v_{i+1}) \in T$.
prefix of a play in $A$.\\
%finite path in $(S,{\rightarrow})$.\\

@@

% \subsubsection{Strategies}
\subsection{Strategies}

% Before deﬁning formally what it means for a player to win a game, we need to introduce the notion of strategy.

@@

%\noindent
%Note that the above definitions are essentially the same for finite and infinite arenas.

 \subsubsection{Winning conditions}

@@

% \paragraph{Reachability Games}
\subsection{Reachability Games}

@@

With regards to reachability games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem. \newline

@@

% \paragraph{Parity Games}
\subsection{Parity Games}

@@

With regards to parity games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem. \newline

@@

The transition graph of the $n$-pebble pushdown automaton $\ppda$ will simulate the possible transitions graphs 
(one for every 
%total 
parameter assignment)
 of the pushdown automaton with $n$ parameters by using pebbles to represent the parameters. 
The $n$-pebble pushdown automaton will first simulate the 
parameter valuation arena.
Players take turns chosing particular stack contents on which to place pebbles,
using a gadget similar as the one from Figure~\ref{quantifiers}. 
Once every parameter has been assigned with the dropping of a corresponding pebble, 
the pebble assignment $\mu$ is fixed.
%%
%
%
%
%
%
%

@@

Concerning higher-order pushdown automata, we are interested in the following decision problem. 
% \newpage

>>>>

%ù \subsection{Games}

A {\em game} is composed of an arena and a winning condition.
We will ﬁrst study
arenas and
 then introduce common winning conditions that will be of interest to us.

% \subsubsection{Arenas}

\begin{samepage}
An {\em arena} is a tuple $A=(S_0, S_1,{\rightarrow})$  which is composed of
\begin{itemize}
	\item two  disjoint  sets of configurations, $S_0$ and $S_1$, %respectively associated with player $0$ and $1$, 
	whose disjoint union $S_0 \cup S_1$ we denote by $S$, and
	\item a relation of %labeled 
		transitions
	 ${\rightarrow} \subseteq S \times %\Lambda \times 
	 				S$. 
\end{itemize}
\end{samepage}

@@

%  \subsubsection{Plays}

\noindent
A {\em play} in an arena $A=(S_0,S_1,\rightarrow)$ is a path in $(S,{\rightarrow})$ that is {\em maximal} in the following sense: it is either infinite or finite and if it is finite, then the target configuration is a dead end. 
A {\em partial play} in an arena $A=(S_0,S_1,\rightarrow)$ is a 
% sequence of configurations $s_{init} = s_0 s_1 \ldots \in S^\infty$ connected by the transition relation, i.e. such that $(v_i, v_{i+1}) \in T$.
prefix of a play in $A$.\\
%finite path in $(S,{\rightarrow})$.\\

@@

% \subsection{Strategies}

@@

%  \subsubsection{Winning conditions}

@@

% \subsection{Reachability Games}

@@

With regards to reachability games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem.

@@

% \subsection{Parity Games}

@@

With regards to parity games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem.

@@

The transition graph of the $n$-pebble pushdown automaton $\ppda$ will simulate the possible transitions graphs 
(one for every 
%total 
parameter assignment)
 of the pushdown automaton with $n$ parameters by using pebbles to represent the parameters. 
The $n$-pebble pushdown automaton will first simulate the 
parameter valuation arena.
Players take turns chosing particular stack contents on which to place pebbles. 
Once every parameter has been assigned with the dropping of a corresponding pebble, 
the pebble assignment $\mu$ is fixed.
%%
%
%
%
%
%
%

@@

Concerning higher-order pushdown automata, we are interested in the following decision problem.
